The Space of Feasible Execution Times for Asynchronous Periodic Task Systems using Definitive Idle Times

### 2013-09-25 dit-paper_ieee.tex
merge Laurent George

\author{
\IEEEauthorblockN{Thomas Chapeaux and Paul Rodriguez}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles\\
tchapeau@ulb.ac.be, paurodri@ulb.ac.be}
\and
\IEEEauthorblockN{Laurent George}
\IEEEauthorblockA{ECE Paris\\
lgeorge@ece.fr}
\and
\IEEEauthorblockN{Jo\"el Goossens}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles\\
joel.goossens@ulb.ac.be}
}

@@

\begin{abstract}
	Sensitivity analysis for real-time systems provides efficient methods to determine feasibility
	conditions in the case of small changes in task parameters, such as
	different WCET values when porting a system to a different platform. The \mbox{C-space} describes the set of execution
	times for which a system is feasible, which allows system designers to quickly
	find the best platform for a given task set. This paper extends previous works
	generalizing the description of the C-space for Earliest Deadline First
	scheduling to systems with initial task offsets. The feasibility gain offered
	by offsets is then expressed as the volume ratio between the C-space of an
	asynchronous system and its corresponding synchronous system. The problem
	requires solving multiple similar instances of the Chinese Remainder Problem,
	for which an algorithm efficient in this context is provided.
\end{abstract}

@@

	Abstract models for real-time systems often require the determination of the
	worst-case execution time of a job (WCET), a value highly dependent on the
	platform on which the application is deployed. Feasibility conditions allow to
	decide whether a system can be scheduled correctly (i.e. without violating any
	timeliness constraint) based on task parameters, including the WCET.

	The \emph{C-space} of a real-time system is, intuitively,
	the set of WCET values for which this system would be feasible
	\cite{bini2004schedulability,george2009characterization}. A precise and efficient description of the C-space of a system would allow to quickly decide
	whether or not it is feasible on a given platform.

	In section~\ref{sct:relatedWorks}, the description of the C-space of synchronous
	constrained systems, which has been mostly covered by \cite{george2009characterization}, is revisited.
	This description is based on the DBF feasibility condition described in
	\cite{baruah1999generalized,baruah1990algorithms} and is found with exponential complexity using
	the notion of definitive idle times (DIT).

@@

		The following necessary and sufficient condition of feasibility
		based on the DBF is given in \cite{baruah1999generalized}:
		\begin{theorem}
			\begin{equation*}
				\begin{array}{c}
					\tau \: \text{is feasible}  \iff
					\dbf{t_1, t_2} \leqslant t_2 - t_1 \; \forall t_1, t_2 \mid  0 \leq t_1 \leq t_2
				\end{array}
			\end{equation*}
		\end{theorem}
        It is easy to see that the condition is at least necessary, as if there exists values of $t_1$ and $t_2$ such that $\dbf{t_1, t_2} > t_2 - t_1$, it means that more than $t_2 - t_1$ units of computation have to be done in $t_2 - t_1$ units of time. Therefore the system would not be feasible.

@@

  \begin{definition}
    A \textbf{definitive idle time} (DIT) \cite{lipariaverage} is a time $t$ such that every job
    released strictly before instant $t$ has its absolute deadline before or at instant $t$.
  \end{definition}

  \begin{definition}
    The earliest DIT occurring in the system strictly after $t=0$ is called the
    \textbf{first DIT} and is noted $t_d$.
  \end{definition}

		\begin{remark} (from the definitions):
			\begin{itemize}
				\item Contrary to other idle times, DITs are independent of the
					scheduling or the execution times of the jobs.
				\item In a feasible system, every DIT is an idle time.
				\item In a feasible system, the first DIT happens after the end of the
					first busy period.
				\item In synchronous systems any $kH$ with $k \geqslant 0$ is a DIT and we
					also have $t_d \leqslant H$.
			\end{itemize}
		\end{remark}

@@

  For synchronous systems in which the execution times are not known, the first DIT $t_d$
  is our ``best guess'' at an upper limit on the end of the busy period. As it was sufficient
  to check the DBF test in intervals $[0, t]$ before the end of the busy period, it is
  slightly more pessimistic but still correct to use the following condition:
  \begin{equation}
  		\dbf{0,t} \leqslant t \; \forall t \leqslant t_d
  		\iff \text{The system is feasible}
  \end{equation}

  To summarize, the description of the C-space is reduced from an infinite number
  of constraints to only those where $t_1 = 0$, $t_2 \leqslant t_d$ and
  values of $t_2$ correspond to job deadlines. Note that the number of
  constraints is still exponential in the number of tasks in the worst-case ($t_d = H$).

@@

		This system is equivalent to multiple instances (one per combination of $a_i$
		values) of the Chinese Remainder Problem (CRP) where the goal is to find the
		value of $t_d$ such that it is congruent to $a_i+O_i$ modulo $T_i$ for all
		tasks $\tau_i$ in $\tau$. To find the FPDIT it is to the best of our knowledge
		necessary to solve the problem on all combinations of $a_i$ values then
		compute the minimum solution. We define $x_i = a_i+O_i$ and $n_i = T_i$ for
		clarity in the context of the CRP. One instance of this problem is guaranteed
		to have at least one solution modulo $H$ for given values of $a_i$ if
		Condition~\ref{eq:CRPcond} is satisfied \cite{knuth1969art}:
		\begin{equation}
			\label{eq:CRPcond}
			x_i \equiv x_j \; (\operatorname{mod} \; \operatorname{gcd}(n_i, n_j))
			\; \forall \tau_i,\tau_j \in \tau
		\end{equation}
		Condition~\ref{eq:CRPcond} means that DITs might still exist even though in
		the general case task periods will not necessarily be pairwise coprime.
% 		We now present an example of such a system.\\
%
% 		\begin{center}
% 			\begin{tabular}{|r|c|c|c|c|}
% 				\hline
% 							& $O_i$ & $C_i$ & $D_i$ & $T_i$ \\ \hline
% 				$\tau_1$ 	& 0 	& $C_1$ & 3 	& 4\\ \hline
% 				$\tau_2$ 	& 2 	& $C_2$ & 7 	& 8\\ \hline
% 			\end{tabular}
% 		\end{center}
%
% 		Here, the possible values of $a_1$ are 3 and 4, and the possible values
% 		of $a_2$ are 7 and 8. For a FPDIT to exist, one of the following
% 		conditions must be fulfilled:
% 		\[
% 			\left[
% 			\begin{array}{c}
% 				3 + 0 \equiv 7 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				3 + 0 \equiv 8 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				4 + 0 \equiv 7 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				4 + 0 \equiv 8 + 2 \; (\operatorname{mod} \; 4 ) \\
% 			\end{array}
% 			\right.
% 		\]
% 		However neither is, which means the system has
% 		no FPDIT (and no other DIT than $t=0$).

	\subsection{Solving the modular system}
		A sketch of the construction algorithm for $t_d$ can be found in
		Algorithm~\ref{alg:algoCRP} (note that $x^{-1} \; (mod \; y)$ means the
		modular inverse of $x$ in the naturals modulo $y$). In this paper we
		concentrate on the latter part of the algorithm (after the modifications to
		correct non-coprime periods, readers interested in the details of these
		modifications should refer to \cite{knuth1969art}).
		\begin{algorithm}
			\caption{Gauss's CRP Algorithm}
			\label{alg:algoCRP}
			\begin{algorithmic}[1]
				\REQUIRE $\forall \tau_i \in \tau : a_i \in [D_i,T_i]$
				\REQUIRE $\forall \tau_i \in \tau : x_i = a_i+O_i$
				\REQUIRE $\forall \tau_i \in \tau : n_i = T_i$
				\REQUIRE $H$ is the hyperperiod of the system

@@

			For this reason, as shown by \cite{leung1982complexity}, it is sufficient (but pessimistic) to
			check for
			intervals included in the interval $[O_{max}, O_{max} + 2H]$, which has an exponential length
			in the number of tasks. A better interval is given in \cite{choquet2004minimal} using the
			notion of cyclic idle times (denoted $t_c$), which are dependent of the execution
			time and therefore cannot be used directly. Their interval is $[0, t_c + H]$ and is valid for
			every cyclic idle time $t_c$.

@@

The general problem of redundancy of some linear constraint $C \, X \leq d$ w.r.t. a set of previous linear constraints $A \, X \leq B$ in an ILP can itself be written as the ILP presented in Fig.~\ref{fig:redILP}.

\begin{figure}[h]
$$z = \max C \, X$$
s.t.
\[
\begin{array}{rccc}
  A & X &\leq & B \\
  C & X &\leq & d + 1
\end{array}
\]
with
\[
  \begin{array}{ccc}
    A & : & [n,k] \text{ matrix}\\
    B & : & [n,1] \text{ matrix}\\
    C & : & [n,1] \text{ matrix}\\
    X & : & [1,n] \text{ matrix}\\
    d & : & \text{integer}
  \end{array}
\]
\caption{Redundancy of $CX \leq d$ w.r.t. $A X \leq B$ as an ILP}
\label{fig:redILP}
\end{figure}

@@

\section{Experimental C-space size gain of asynchronous systems}
\label{sct:expCspaceGain}
	One of the advantages of asynchronous systems over synchronous systems in the
	context of periodic tasks is that the critical instants of asynchronous systems
	are at most as constraining for the C-space as a synchronous instant. For this
	reason, modifying task offsets when possible during
	the design of a real-time system is a useful trick to fit a heavy periodic task set
	into a given platform.

@@

	In other words, fig.~\ref{fig:sizeRatio}
	evaluates the feasibility gain of adding random offsets to a synchronous task
	set. It is assumed that a more advanced strategy to assign these offsets (such
	as the ones from \cite{goossens2003scheduling}) would yield a
	greater difference.

@@

	The ratio between the size of the synchronous and asynchronous C-spaces of
	these systems roughly range from 0.95 in the worst case (high utilization, high
	CDF) to 0.85 in the best case (low utilization, low CDF). As calculating those
	sizes is rather long, the number of systems used to obtain average
	results was rather low (5000), which is the likely cause of the spikes in the
	data. Note that the value of the CDF has a greater influence on this ratio than
	the system's total utilization, but less so when the CDF is lower. It is also
	important to note that even though a 0.95 ratio may not seem much, the
	synchronous C-spaces of the systems contain many dominated WCET vectors, and
	only those that are in the border (i.e. non-dominated) actually matter to the
	designer of the system. Using the asynchronous C-space instead will relax some
	constraints and therefore add new acceptable solutions that will dominate solutions that were previously non-dominated,
	increasing the quality of the C-space border.

@@

Future works could also repeat the simulation of section~\ref{sct:expCspaceGain} with the algorithm described in \cite{grenier2008pushing}, which determine offset values with optimal feasibility. The
feasibility gain could then be expressed using the C-space.

>>>>

\author{
\IEEEauthorblockN{Thomas Chapeaux and Paul Rodriguez}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles / ECE\\
tchapeau@ulb.ac.be, paurodri@ulb.ac.be}
\and
\IEEEauthorblockN{Laurent George}
\IEEEauthorblockA{University of Paris-Est / ECE Paris\\
lgeorge@ieee.org}
\and
\IEEEauthorblockN{Jo\"el Goossens}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles\\
joel.goossens@ulb.ac.be}
}

@@

\begin{abstract}
	Sensitivity analysis for real-time systems provides efficient methods to determine feasibility
	conditions in the case of changes in task parameters, such as
	different WCET values when porting a system to a different platform. The \mbox{C-space} defines the space of execution
	times for which a system is feasible, which can help system designers to more easily check 
	if a task set is feasible on a set of candidate platforms. This paper extends previous works
	generalizing the description of the C-space for Earliest Deadline First
	scheduling to systems with initial task offsets. The feasibility gain offered
	by offsets is then expressed as the volume ratio between the C-space of an
	asynchronous system and its corresponding synchronous system when offsets are randomly chosen. The problem
	requires solving multiple similar instances of the Chinese Remainder Problem,
	for which an efficient algorithm in this context is provided. We show that even for random offsets, the gain can reach 15%.
\end{abstract}

@@

	Abstract models for real-time systems often require the determination of the
	worst-case execution time of a job (WCET), a value highly dependent on the
	platform on which the application is deployed. Feasibility conditions allow us to
	decide whether a system can be correctly scheduled (i.e. without violating any
	timeliness constraint) based on task parameters, including the WCET.

	The \emph{C-space} of a real-time system is, intuitively,
	the space of WCET values for which this system is feasible
	\cite{bini2004schedulability,george2009characterization}. A precise and efficient description of the C-space of a system would allow us to efficiently decide whether or not it is feasible on a set of candidate platforms.

	In section~\ref{sct:relatedWorks}, the description of the C-space of synchronous
	constrained systems, which has been mostly covered by \cite{george2009characterization}, is revisited.
	This description is based on the Demand Bound Function (DBF) feasibility condition described in
	\cite{baruah1999generalized,baruah1990algorithms} and is found with exponential complexity using
	the notion of definitive idle times (DIT). The DIT notion provides a feasibility interval for EDF that does not depend of the WCET of the tasks.

@@

		The following theorem is a necessary and sufficient condition of feasibility
		based on the DBF, given in \cite{baruah1999generalized}:
		\begin{theorem}
			\begin{equation*}
				\begin{array}{c}
					\tau \: \text{is feasible}  \iff
					\dbf{t_1, t_2} \leqslant t_2 - t_1 \; \forall t_1, t_2 \mid  0 \leq t_1 \leq t_2
				\end{array}
			\end{equation*}
		\end{theorem}
        %It is easy to see that the condition is at least necessary, as if there exists values of $t_1$ and $t_2$ such that $\dbf{t_1, t_2} > t_2 - t_1$, it means that more than $t_2 - t_1$ units of computation have to be done in $t_2 - t_1$ units of time. Therefore the system would not be feasible.

@@

  \begin{definition}
    A \textbf{definitive idle time} (DIT) \cite{lipariaverage} is a time $t$ (if it exists) such that every job
    released strictly before instant $t$ has its absolute deadline before or at instant $t$.
  \end{definition}

  \begin{definition}
    The earliest DIT (if it exists) occurring in the system strictly after $t=0$ is called the
    \textbf{first DIT} and is noted $t_d$.
  \end{definition}

		\begin{remark} (from the definitions, if a DIT exist then):
			\begin{itemize}
				\item Contrary to other idle times, a DIT is independent of the
					scheduling or the execution times of the jobs.
				\item In a feasible system, every DIT is an idle time.
				\item In a feasible system, the first DIT happens after the end of the
					first busy period.
				\item In synchronous systems any $kH$ with $k \geqslant 0$ is a DIT and we
					also have $t_d \leqslant H$.
			\end{itemize}
		\end{remark}

@@

  For synchronous systems in which the execution times are not known, the first DIT $t_d$
  is our ``best guess'' at an upper limit on the end of the busy period. As it was sufficient
  to check the DBF test in intervals $[0, t]$ before the end of the first busy period. This time interval is WCETs dependant. A more general feasility condition valid for any WCET is as follows:
  \begin{equation}
  		\dbf{0,t} \leqslant t \; \forall t \leqslant t_d
  		\iff \text{The system is feasible}
  \end{equation}

  To summarize, the description of the C-space is reduced from an infinite number
  of constraints to only those where $t_1 = 0$, $t_2 \leqslant t_d$ and
  values of $t_2$ correspond to job deadlines. Note that the number of
  constraints is still exponential in the number of tasks in the worst-case (which is when $t_d = H$).

@@

		This system is equivalent to multiple instances (one per combination of $a_i$
		values) of the Chinese Remainder Problem (CRP) where the goal is to find the
		value of $t_d$ such that it is congruent to $a_i+O_i$ modulo $T_i$ for all
		tasks $\tau_i$ in $\tau$. To find the FPDIT it is to the best of our knowledge
		necessary to solve the problem on all combinations of $a_i$ values and then
		to compute the minimum solution. We define $x_i = a_i+O_i$ and $n_i = T_i$ for
		clarity in the context of the CRP. One instance of this problem is guaranteed
		to have at least one solution modulo $H$ for given values of $a_i$ if
		Condition~\ref{eq:CRPcond} is satisfied \cite{knuth1969art}:
		\begin{equation}
			\label{eq:CRPcond}
			x_i \equiv x_j \; (\operatorname{mod} \; \operatorname{gcd}(n_i, n_j))
			\; \forall \tau_i,\tau_j \in \tau
		\end{equation}
		Condition~\ref{eq:CRPcond} means that DITs might still exist even though in
		the general case task periods will not necessarily be pairwise coprime.
% 		We now present an example of such a system.\\
%
% 		\begin{center}
% 			\begin{tabular}{|r|c|c|c|c|}
% 				\hline
% 							& $O_i$ & $C_i$ & $D_i$ & $T_i$ \\ \hline
% 				$\tau_1$ 	& 0 	& $C_1$ & 3 	& 4\\ \hline
% 				$\tau_2$ 	& 2 	& $C_2$ & 7 	& 8\\ \hline
% 			\end{tabular}
% 		\end{center}
%
% 		Here, the possible values of $a_1$ are 3 and 4, and the possible values
% 		of $a_2$ are 7 and 8. For a FPDIT to exist, one of the following
% 		conditions must be fulfilled:
% 		\[
% 			\left[
% 			\begin{array}{c}
% 				3 + 0 \equiv 7 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				3 + 0 \equiv 8 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				4 + 0 \equiv 7 + 2 \; (\operatorname{mod} \; 4 ) \\
% 				4 + 0 \equiv 8 + 2 \; (\operatorname{mod} \; 4 ) \\
% 			\end{array}
% 			\right.
% 		\]
% 		However neither is, which means the system has
% 		no FPDIT (and no other DIT than $t=0$).

	\subsection{Solving the modular system}
		A sketch of the construction algorithm to compute $t_d$ can be found in
		Algorithm~\ref{alg:algoCRP} (note that $x^{-1} \; (mod \; y)$ means the
		modular inverse of $x$ in the naturals modulo $y$). In this paper we
		concentrate on the latter part of the algorithm (after the modifications to
		correct non-coprime periods, readers interested in the details of these
		modifications should refer to \cite{knuth1969art}).
		\begin{algorithm}
			\caption{Gauss's CRP Algorithm}
			\label{alg:algoCRP}
			\begin{algorithmic}[1]
				\REQUIRE $\forall \tau_i \in \tau : a_i \in [D_i,T_i]$
				\REQUIRE $\forall \tau_i \in \tau : x_i = a_i+O_i$
				\REQUIRE $\forall \tau_i \in \tau : n_i = T_i$
				\REQUIRE $H$ is the hyperperiod of the system

@@

			For this reason, as shown by \cite{leung1982complexity}, it is sufficient (but pessimistic) to
			check for
			intervals included in the interval $[O_{max}, O_{max} + 2H]$, which has an exponential length
			in the number of tasks. A better interval is given in \cite{choquet2004minimal} using the
			notion of cyclic idle times (denoted $t_c$), which are dependent of the execution
			time and therefore cannot be used directly to characterize the C-Space. Their interval is $[0, t_c + H]$ and is valid for
			every cyclic idle time $t_c$.

@@

The general problem of redundancy of some linear constraint $C \cdot X \leq d$
w.r.t. a set of previous linear constraints $A \cdot X \leq B$ in an ILP can
itself be written as the ILP presented in Fig.~\ref{fig:redILP}.

\begin{figure}[h]
$$z = \max C \cdot X$$
s.t.
\[
\begin{array}{rcc}
  A \cdot X &\leq & B \\
  C \cdot X &\leq & d + 1
\end{array}
%\begin{array}{rccc}
%  A & X &\leq & B \\
%  C & X &\leq & d + 1
%\end{array}
\]
with
\[
  \begin{array}{ccc}
    A & : & [n,k] \text{ matrix}\\
    B & : & [n,1] \text{ matrix}\\
    C & : & [n,1] \text{ matrix}\\
    X & : & [1,n] \text{ matrix}\\
    d & : & \text{integer}
  \end{array}
\]
\caption{Redundancy of $C \cdot X \leq d$ w.r.t. $A \cdot X \leq B$ as an ILP}
\label{fig:redILP}
\end{figure}

@@

\section{Experimental C-space size gain of asynchronous systems}
\label{sct:expCspaceGain}
	One of the advantages of asynchronous systems over synchronous systems in the
	context of periodic tasks is that the critical instants of asynchronous systems
	can be less constraining for the C-space than synchronous instants (if offsets are chosen to prevent synchronous releases). For this reason, modifying task offsets when possible during
	the design of a real-time system is a useful trick to fit a heavy periodic task set
	into a given platform.

@@

	In other words, fig.~\ref{fig:sizeRatio}
	evaluates the feasibility gain of adding random offsets w.r.t. a synchronous task
	set. We have chosen in this paper to generate offsets randomly but we did not consider offsets that eventually lead to synchronous releases. The problem of offset assignment has been considered for example in \cite{grenier2008pushing}. The problem of those heuristics is that they depend on WCETs. Finding offsets that maximize a C-space region is still an open problem.

@@

	The ratio between the size of the synchronous and asynchronous C-spaces of
	these systems roughly range from 0.95 in the worst case (high utilization, high
	CDF) to 0.85 in the best case (low utilization, low CDF). Each point on the graph is based on an average over 5000 tests.
	Note that the value of the CDF has a greater influence on this ratio than
	the system's total utilization, but less so when the CDF is higher. It is also
	important to note that even though a 0.95 ratio may not seem much, the
	synchronous C-spaces of the systems contain many dominated WCET vectors, and
	only those that are in the border (i.e. non-dominated) actually matter to the
	designer of the system. Using the asynchronous C-space instead will relax some
	constraints and therefore add new acceptable solutions that will dominate solutions that were previously non-dominated,
	increasing the quality of the C-space border.

@@

Future works could also repeat the simulation of section~\ref{sct:expCspaceGain} with the algorithm described in \cite{grenier2008pushing}, which determine offset values with optimal feasibility but for specific WCETs. We would like to study if those heuristics obtained for particular WCETs scenarios can be also interesting for any WCET in the C-Space.
